Simulation of angular resolution of a new electromagnetic sampling calorimeter

### 2023-01-26 paper.tex
uodates

\begin{abstract}
We report on the simulation results for the angular resolution of an electromagnetic (EM) sampling calorimeter with photons in the range of 100~MeV to 2~GeV. The simulation model of the EM calorimeter consists of alternating layers of an 1-mm-thick lead plate and a 5-mm-thick plastic scintillator plate. The scintillator plates are alternately segmented into horizontal and vertical strips. In this study, we obtain energy deposits in individual strips using Geant4 simulations and reconstruct the incident photon angles using XGBoost with gradient-boosted decision trees. The performance of the angle reconstruction depends on both the detector configuration and accuracy of machine learning. The angular resolution so obtained can be expressed as $0.24 \oplus 1.25/\sqrt{E_{\gamma}}$, where $E_{\gamma}$ is the incident photon energy in GeV, for strips of 15 mm and 32 layers. This energy dependence is consistent for different incident angles in the range of 10$^{\circ}$ to 40$^{\circ}$.

@@

The sampling calorietmer has been used to measure the direction of arriving photons in a space-based detector to identify astrophysical sources for gamma-ray astrophysics. The Fermi Large Area Telescope (LAT) has a tracker consisting of 18 $x$-$y$ pairs of silicon strip detectors with 16 interleaved tungsten foils. Tracking with the silicon detectors on the electron-positron pairs produced at the foils provides an angular resolution of 0.6 degrees for 1-GeV photons and better than 0.16 degrees for photons with energies higher than 10~GeV~\cite{FERMI:LAT}. The LAT has two additional components: a plastic scintillator for the charged-particle detection and a homogeneous CsI crystals for the energy measurement. This triple detector configuration is also used in different missions. An energy spectrum measurement of cosmic-ray electrons (CALET)~\cite{CALET} utilizes scintillating fibers and PWO crystals instead of the silicon strips and the CsI crystals, respectively. A dark matter particle explorer mission (DMPE)~\cite{DMPE} selects BGO crystals for the energy measurement. The DMPE provides the angular resolution of 1--2 degrees for 1-GeV photons.

@@

The $K_{L} \rightarrow \pi^{0}\nu\bar{\nu}$ decay is identified as a final state existing only a single $\pi^{0}$ reconstructed from detected two photons. A hermetic detector system enclosing the decay region should be prepared to confirm only two photons in the final state. There had been an experimental proposal to detect the $K_{L} \rightarrow \pi^{0}\nu\bar{\nu}$ decay with the reconstruction of a single $\pi^{0}$ using directional information of two photons~\cite{KOPIO}. A Preradiator (PR) is placed in front of the calorimeter to measure the photon's incident angle. The PR consists of a close-packed multi-layer sandwich of 8-mm-thick plastic scintillators alternating with the single-plane and this-gap drift chambers with cathode strips. For good energy resolution and high detection efficiency against a few hundred MeV photons, the plastic scintillator is used for electron-positron conversion. This idea had been confirmed by a beam test using tagged photon beams and achieved its angular resolution of 25~mrad for 250-MeV~photons~\cite{KOPIOcdr}.

On the other hand, the KOTO experiment at J-PARC~\cite{KOTOexp} takes another approach to identify $K_{L} \rightarrow \pi^{0}\nu\bar{\nu}$ decay with a well-collimated small-size neutral beam, which provides a condition that the $\pi^{0}$ decays at the beam axis. With this assumption, the decay vertex is calculated using the positions and energies of the two photons measured in the calorimeter. The transverse momentum of $\pi^{0}$ is calculated from the vertex. The two variables, the decay vertex and the transverse momentum are used to define a signal region of the decay. The KOTO experiment started data taking since 2013 and has achieved its experimental sensitivity as 7.2~$\times$~10$^{-10}$~\cite{KOTOexp}. The data taking will continue for a few more years, and the sensitivity will reach 10$^{-11}$.

@@

\begin{figure}[!hbt]
\centering
\includegraphics[width=0.6\textwidth]{Fig1_detector_schematic.jpeg}
\caption{ Schematic of the sampling calorimeter model consisting of 105 alternating layers of lead and scintillator plates. Each plate is segmented in 35 strips, oriented alternately in horizontal and vertical directions. See text for details. }
\label{fig:det_conf}
\end{figure}

The toy sampling calorimeter is designed as a block consisting of alternating layers of an 1-mm-thick lead absorber and a 5-mm-thick polyvinyltoluene-based plastic scintillator. The plastic scintillator is segmented into 15-mm-wide strips, which are alternately oriented in the vertical and horizontal directions, as shown in Figure~\ref{fig:det_conf}. It has a cross-section size of 525~$\times$~525~mm$^{2}$ and accommodates 105 alternating layers of 630~mm and 20 radiation lengths (20$X_{0}$) to contain photons showers for energies in the range of 0.1 to 2~GeV.

@@

Training data for machine learning are prepared using the Geant4 simulation such that the input dataset is representative of the detector response of the sampling calorimeter. In order to minimize the bias in the training process related to the dataset, the incident angles are uniformly generated at the detector surface in the range of 0~$<\theta<$~50$^{\circ}$ and 0~$<\varphi<$~360$^{\circ}$, where $\varphi$ denotes the azimuthal angle. The number of training samples is $10^{5}$ considering the limited computing resources. To test the reconstruction of the incident angles, photons are generated at a fixed incident angle $\theta$ with known incident energy and incident position. We do not take into account any arbitrary noises that would affect the angle reconstruction. Instead, the angle reconstruction is tested with a lower threshold of 0.5~MeV for each strip, and the angular resolution does not change.

@@

The incident angle of the photons is reconstructed using XGBoost~\cite{xgboost:2016}, which maps a feature dataset, the energy deposit in each scintillator strip, onto a target variable, the incident angle of a photon. The feature size identical to the number of strips varies with the strip width to have the same coverage. In the case of the 15-mm-wide strips, 35~$\times$~105 channels compose the detector. XGBoost constructs gradient-boosted decision trees using the training dataset, efficiently descending the loss function. The hyperparameter of XGBoost is the parameter driving how models are trained. In this paper, main five parameters are studied. \textit{N\_estimators} defines the maximum number of allowed decision trees to be developed, and \textit{Max. depth} defines the complexity of the decision-tree structure. \textit{Subsample} controls the fraction of total event samples for each boosting procedure, \textit{learning rate} weights a decision tree to be added onto the current model, and \textit{gamma} regulates the evaluation of each decision tree.

@@

\begin{figure}[!hbt]
\centering
\includegraphics[width=0.79\textwidth]{Fig4_include_ne250.pdf}
\caption{Angular resolutions are displayed in terms of combination of \textit{N\_estimators} and \textit{Max. depth}. The best angular resolution is obtained with \textit{N\_estimators}~=~300 and \textit{Max. depth}~=~100. }
\label{fig:par_scan}
\end{figure}

Because the performance of the angle reconstruction of the XGB depends on the hyperparameters, we scan the angular resolution by changing them. Since there are correlations among the hyperparameters, the scanning process is performed in a five-dimensional space for all possible combinations. As an example, Figure~\ref{fig:par_scan} shows the test results for \textit{N\_estimators} and \textit{Max. depth}. We explore a hyperparameter combination that generates the best angular resolution within the uncertainty of 0.01~deg. Consequently, the \textit{N\_estimators} and \textit{Max. depth} are set to 300 and 100, respectively. Similar processes are also performed for different hyperparameters, \textit{subsample}, \textit{learning rates}, and \textit{gamma}. Table~\ref{tab:XgbPar} lists the optimized set of hyperparameters that are used in further studies. During the optimization process, the fraction of the tail that is not well described by the GG function is approximately 2\% for different sets of hyperparameters.

@@

\begin{figure}[!hbt]
\centering
\includegraphics[width=0.58\textwidth]{Fig5_width_including7.5_12.5mm.pdf}
\caption{ Angular resolutions are deduced in terms of the strip width for 1-GeV photons at $\theta=$~10$^{\circ}$ }
\label{fig:angle_reco_width}
\end{figure}
\subsubsection{Strip width}
To inspect how much the angular resolution is affected by the spatial resolution of measured EM showers, Angular resolutions are deduced by varying the strip width from 5 to 35 mm, as shown in Figure~\ref{fig:angle_reco_width}. The narrower the strip width, the larger the feature size. Consequently, the angle-reconstruction performance is better with 10-mm-wide strips than with 5-mm-wide strips. However, the wider strips hardly provide detailed information on the EM shower. The 15-mm-wide strips yield a good angular resolution of 1.23~$\pm$~0.01$^{\circ}$ and are selected for further study considering the actual construction of a prototype detector.

@@

\subsubsection{Number of layers}
To further study the feasibility for the detector to be a photon tracker, the reconstruction of the incident angle is performed with the front layers instead of the full layers. Figure~\ref{fig:angle_reco_layer} (a) shows the reconstructed angular distribution for 1-GeV photons by using only the front 32 layers, which correspond to 6.2$X_{0}$. A fraction of the photons fails to be reconstructed because of insufficient information gathering caused by deeply penetrating photons without shower generation, where the fraction is estimated to be 1.9\%. The failed events are represented as a delta function near 0. Such events are removed by requiring the total energy deposit at the front layers to be higher than 10~MeV, which is 1\% of the incident energy. The angular resolution with the front 32 layers is estimated to be 1.30~$\pm$~0.01$^{\circ}$. The evolution of angular resolution degrades with the reduction of the number of layers used as the number of contributors to the reconstruction decreases.

@@

Figure~\ref{fig:res_edep} shows the estimated $p_{0}$ and $p_{1}$ as a function of the incident angle. The average values of $p_{0}$ and $p_{1}$ are estimated to be 0.24$^{\circ}$ and 1.25$^{\circ}$, respectively, for $\theta>$~10$^{\circ}$. In the case of $\theta$ smaller than 10 degrees, the angular resolution shows different dependency on the incident energy, which results in smaller $p_{1}$ and larger $p_{0}$ compared to the larger angles.

@@

\subsubsection{Training sample}
Figure~\ref{fig:multi-parameter} shows the angular resolution for different numbers of training samples and detector configurations. The width of each curve represents the statistical uncertainty. All configurations show improvement in the angular resolution with the increasing number of training samples, which indicates that the maturity of machine learning is an important factor in the angle reconstruction. With an increase in the number of training events, the angular resolution improves most rapidly for the configuration with 5-mm-wide scintillator strips. In other words, the required number of training events should be large enough to provide correct answer from the correlations between many features.

@@

A sampling calorimeter for the incident angle measurement of photons will play an essential role in the next stage of the KOTO experiment, which can distinguish $\pi^0$ decays at the off- from in-beam axis. The machine-learning approach (XGboost) analyzing the energy-weighted shower shapes enables us to proceed with detailed optimization of the detector to fulfill the experimental goal. The angular resolution depends on the values of hyperparameters of the XGboost decided by the user.

@@

For the toy sampling calorimeter with 15-mm-wide strips and 32 layers, the angular resolution was obtained as 1.30~$\pm$~0.01$^{\circ}$ in machine learning with $10^{5}$ training samples. Energy dependence of the angular resolution can be expressed as $p_{0} \oplus p_{1}/\sqrt{E_{\gamma}}$. For the angular range of $10^{\circ}<\theta<40^{\circ}$, the  $p_{0}$ and $p_{1}$ values remain unchanged as $p_{0}$ = 0.238$^{\circ}$ $\pm$ 0.012$^{\circ}$ and $p_{1}$ = 1.248$^{\circ}$ $\pm$ 0.002$^{\circ}$, respectively.

>>>>

\begin{abstract}
We report on the simulation results for the angular resolution of an electromagnetic (EM) sampling calorimeter with photons in the range of 100~MeV to 2~GeV. The simulation model of the EM calorimeter consists of alternating layers of a 1-mm-thick lead plate and a 5-mm-thick plastic scintillator plate. The scintillator plates are alternately segmented into horizontal and vertical strips. In this study, we obtain energy deposits in individual strips using Geant4 simulations and reconstruct the incident photon angles using XGBoost with gradient-boosted decision trees. The performance of the angle reconstruction depends on the detector configuration and the accuracy of machine learning. The angular resolution is well described by the expression $0.24^{\circ} \oplus 1.25^{\circ}/\sqrt{E_{\gamma}}$, where $E_{\gamma}$ is the incident photon energy in GeV, for strips of 15 mm and 32 layers. This energy dependence is consistent for different incident angles in the range of 10$^{\circ}$ to 40$^{\circ}$.

@@

The sampling calorietmer has been used to measure the direction of arriving photons in a space-based detector to identify astrophysical sources for gamma-ray astrophysics. The Fermi Large Area Telescope (LAT) has a tracker consisting of 18 $x$-$y$ pairs of silicon strip detectors with 16 interleaved tungsten foils. Tracking with the silicon detectors on the electron-positron pairs produced at the foils provides an angular resolution of 0.6 degrees for 1-GeV photons and better than 0.16 degrees for photons with energies higher than 10~GeV~\cite{FERMI:LAT}. The LAT has two additional components: a plastic scintillator for the charged-particle detection and a homogeneous CsI crystals for the energy measurement. This triple detector configuration is also used in different missions. An energy spectrum measurement of cosmic-ray electrons (CALET)~\cite{CALET} utilizes scintillating fibers and PWO crystals instead of the silicon strips and the CsI crystals, respectively. A dark matter particle explorer mission (DMPE)~\cite{DMPE} selects BGO crystals for the energy measurement. Theses detectors provide the angular resolution of 1-2 degrees for 1-GeV photons.

@@

The $K_{L} \rightarrow \pi^{0}\nu\bar{\nu}$ decay is identified as a final state existing only a single $\pi^{0}$ reconstructed from detected two photons. A hermetic detector system enclosing the decay region should be prepared to confirm only two photons in the final state. There had been an experimental proposal to detect the $K_{L} \rightarrow \pi^{0}\nu\bar{\nu}$ decay with the reconstruction of a single $\pi^{0}$ using directional information of two photons~\cite{KOPIO}. A Preradiator (PR) is placed in front of the calorimeter to measure the photon's incident angle. The PR consists of a close-packed multi-layer sandwich of 8-mm-thick plastic scintillators alternating with the single-plane and thin-gap drift chambers with cathode strips. For good energy resolution and high detection efficiency against a few hundred MeV photons, the plastic scintillator is used for electron-positron conversion. This idea had been confirmed by a beam test using tagged photon beams and achieved its angular resolution of 25~mrad for 250-MeV~photons~\cite{KOPIOcdr}.

On the other hand, the KOTO experiment at J-PARC~\cite{KOTOexp} takes another approach to identify $K_{L} \rightarrow \pi^{0}\nu\bar{\nu}$ decay with a well-collimated small-size neutral beam, which provides a condition that the $\pi^{0}$ decays at the beam axis. With this assumption, the decay vertex is calculated using the positions and energies of the two photons measured in the calorimeter. The transverse momentum of $\pi^{0}$ is calculated from the vertex. The two variables, the decay vertex and the transverse momentum are used to define a signal region of the decay. The KOTO experiment started data taking since 2013 and has achieved its experimental sensitivity as 7.2~$\times$~10$^{-10}$~\cite{KOTOexp}. The data taking will continue for a few more years, and the sensitivity will be expected to reach $\mathcal{O}$(10$^{-11}$).

@@

\begin{figure}[!hbt]
\centering
\includegraphics[width=0.6\textwidth]{Fig1_detector_schematic.jpeg}
\caption{ Schematic of the sampling calorimeter model consisting of 105 alternating layers of lead and scintillator plates. Each plate is segmented in 35 strips, oriented alternately in horizontal and vertical directions. }
\label{fig:det_conf}
\end{figure}

The toy sampling calorimeter is designed as a block consisting of alternating layers of an 1-mm-thick lead absorber and a 5-mm-thick polyvinyltoluene-based plastic scintillator. The plastic scintillator is segmented into 15-mm-wide strips, which are alternately oriented in the vertical and horizontal directions, as shown in Figure~\ref{fig:det_conf}. It has a cross-section size of 525~$\times$~525~mm$^{2}$ and accommodates 105 alternating layers of 630~mm and 20 radiation lengths (20$X_{0}$) to contain the EM showers for energies in the range of 0.1 to 2~GeV.

@@

Training data for machine learning are prepared using the Geant4 simulation such that the input dataset is representative of the detector response of the sampling calorimeter. In order to minimize the bias in the training process related to the dataset, the incident angles are uniformly generated at the detector surface in the range of 0~$<\theta<$~50$^{\circ}$ and 0~$<\varphi<$~360$^{\circ}$, where $\varphi$ denotes the azimuthal angle. The number of training samples is $10^{5}$ considering the limited computing resources. The angle reconstruction is studied using photons generated at given incident $\theta$ angle, energy, and position. It turns out unchanged with the energy deposit threshold above the noise level of 0.5 MeV. Therefore, we do not consider random noises in this study.

@@

The incident angle of the photons is reconstructed using XGB~\cite{xgboost:2016}, which maps a feature dataset, the energy deposit in each scintillator strip, onto a target variable, the incident angle of a photon. The feature size identical to the number of strips varies with the strip width to have the same coverage. In the case of the 15-mm-wide strips, 35~$\times$~105 channels compose the detector. XGB constructs gradient-boosted decision trees using the training dataset, efficiently descending the loss function. The hyperparameter of XGB is the parameter driving how models are trained. In this paper, main five parameters are studied. \textit{N\_estimators} defines the maximum number of allowed decision trees to be developed, and \textit{Max. depth} defines the complexity of the decision-tree structure. \textit{Subsample} controls the fraction of total event samples for each boosting procedure, \textit{learning rate} weights a decision tree to be added onto the current model, and \textit{gamma} regulates the evaluation of each decision tree.

@@

\begin{figure}[!hbt]
\centering
\includegraphics[width=0.79\textwidth]{Fig4_include_ne250.pdf}
\caption{Angular resolutions are displayed in terms of combination of \textit{N\_estimators} and \textit{Max. depth}. The uncertainty for each value is 0.01~deg. }
\label{fig:par_scan}
\end{figure}

Because the performance of the angle reconstruction of the XGB depends on the hyperparameters, we scan the angular resolution by changing them. Since there are correlations among the hyperparameters, the scanning process is performed in a five-dimensional space for all possible combinations. As an example, Figure~\ref{fig:par_scan} shows the test results for \textit{N\_estimators} and \textit{Max. depth}. The angular resolution improves when both \textit{Max. depth} and \textit{N\_estimators} increase. Since the uncertainty of the obtained angular resolution is estimated to be 0.01 degrees, many candidates for the hyperparameter set exist. Considering the CPU time for the training, the \textit{N\_estimators} and \textit{Max. depth} are set to 300 and 100, respectively. Similar processes are also performed for different hyperparameters, \textit{subsample}, \textit{learning rates}, and \textit{gamma}. Table~\ref{tab:XgbPar} lists the optimized set of hyperparameters that are used in further studies. During the optimization process, the fraction of the tail that is not well described by the GG function is approximately 2\%, and the fraction does not change for different sets of hyperparameters.

@@

\begin{figure}[!hbt]
\centering
\includegraphics[width=0.58\textwidth]{Fig5_width_including7.5_12.5mm.pdf}
\caption{ Angular resolutions are deduced in terms of the strip width for 1-GeV photons at $\theta=$~10$^{\circ}$ }
\label{fig:angle_reco_width}
\end{figure}
\subsubsection{Strip width}
To inspect the dependency of the angular resolution on the strip width, we deduce the angular resolution by varying the strip width from 5 to 35mm, as shown in Figure~\ref{fig:angle_reco_width}. The narrower the strip width, the larger the feature size. Consequently, the angle-reconstruction performance is better with 10-mm-wide strips than with 5-mm-wide strips. However, the wider strips hardly provide detailed information on the EM shower. The 15-mm-wide strips yield a good angular resolution of 1.24~$\pm$~0.01$^{\circ}$ and are selected for further study considering the actual construction of a prototype detector, although 10-mm-wide strips provide the best resolution.

@@

\subsubsection{Number of layers}
We study the dependency of the angle reconstruction on the number of front layers. Figure~\ref{fig:angle_reco_layer} (a) shows the reconstructed angular distribution for 1-GeV photons by using only the front 32 layers, which correspond to 6.2$X_{0}$. A fraction of the photons fails to be reconstructed because of insufficient information gathering caused by deeply penetrating photons without shower generation, where the fraction is estimated to be 1.9\%. The failed events are represented as a delta function near 0. Such events are removed by requiring the total energy deposit at the front layers to be higher than 10~MeV, which is 1\% of the incident energy. The angular resolution with the front 32 layers is estimated to be 1.30~$\pm$~0.01$^{\circ}$. The evolution of angular resolution degrades with the reduction of the number of layers used.

@@

Figure~\ref{fig:res_edep} shows the estimated $p_{0}$ and $p_{1}$ as a function of the incident angle. The averages of $p_{0}$ and $p_{1}$ are estimated to be 0.238 $\pm$ 0.012$^{\circ}$ and 1.248 $\pm$ 0.002$^{\circ}$, respectively, for $\theta>$~10$^{\circ}$. In the case of $\theta$ smaller than 10 degrees, the angular resolution shows different dependency on the incident energy, which results in smaller $p_{1}$ and larger $p_{0}$ compared to the larger angles.

@@

\subsubsection{Training sample}
Figure~\ref{fig:multi-parameter} shows the angular resolution for different numbers of training samples and detector configurations. The width of each curve represents the statistical uncertainty. All configurations show improvement in the angular resolution with the increasing number of training samples. With an increase in the number of training events, the angular resolution improves most rapidly for the configuration with 5-mm-wide scintillator strips. In other words, the required number of training events should be large enough to provide correct answer from the correlations between many features.

@@

A sampling calorimeter for the incident angle measurement of photons will play an essential role in the next stage of the KOTO experiment, which can distinguish $\pi^0$ decays at the off- from in-beam axis. The machine-learning approach (XGBoost) analyzing the energy-weighted shower shapes enables us to proceed with detailed optimization of the detector to fulfill the experimental goal. The angular resolution depends on the values of hyperparameters of the XGBoost decided by the user.

@@

For the toy sampling calorimeter with 15-mm-wide strips and 32 layers, the angular resolution was obtained as 1.30~$\pm$~0.01$^{\circ}$ in machine learning with $10^{5}$ training samples. Energy dependence of the angular resolution can be expressed as $p_{0} \oplus p_{1}/\sqrt{E_{\gamma}}$. For the angular range of $10^{\circ}<\theta<40^{\circ}$, the $p_{0}$ and $p_{1}$ remain unchanged as $p_{0}$ = 0.238 $\pm$ 0.012$^{\circ}$ and $p_{1}$ = 1.248 $\pm$ 0.002$^{\circ}$, respectively.
